---
title: "The Analytics Edge Week 1"
author: "Joona Rauhamäki"
date: "15. huhtikuuta 2016"
output: html_document
---

# The Analytics Edge Week 1

Let's start by emptying the workspace:
```{r}
rm (list = ls(all=T))

``` 

## Assignment I An Analytical Detective

Crime is an international concern, but it is documented and handled in very different ways in different countries. In the United States, violent crimes and property crimes are recorded by the Federal Bureau of Investigation (FBI).  Additionally, each city documents crime, and some cities release data regarding crime rates. The city of Chicago, Illinois releases crime data from 2001 onward online.

Chicago is the third most populous city in the United States, with a population of over 2.7 million people. The city of Chicago is shown in the map below, with the state of Illinois highlighted in red. 

There are two main types of crimes: violent crimes, and property crimes. In this problem, we'll focus on one specific type of property crime, called "motor vehicle theft" (sometimes referred to as grand theft auto). This is the act of stealing, or attempting to steal, a car. In this problem, we'll use some basic data analysis in R to understand the motor vehicle thefts in Chicago. 

Please download the file mvtWeek1.csv for this problem (do not open this file in any spreadsheet software before completing this problem because it might change the format of the Date field). Here is a list of descriptions of the variables:

ID: a unique identifier for each observation
Date: the date the crime occurred
LocationDescription: the location where the crime occurred
Arrest: whether or not an arrest was made for the crime (TRUE if an arrest was made, and FALSE if an arrest was not made)
Domestic: whether or not the crime was a domestic crime, meaning that it was committed against a family member (TRUE if it was domestic, and FALSE if it was not domestic)
Beat: the area, or "beat" in which the crime occurred. This is the smallest regional division defined by the Chicago police department.
District: the police district in which the crime occured. Each district is composed of many beats, and are defined by the Chicago Police Department.
CommunityArea: the community area in which the crime occurred. Since the 1920s, Chicago has been divided into what are called "community areas", of which there are now 77. The community areas were devised in an attempt to create socially homogeneous regions.
Year: the year in which the crime occurred.
Latitude: the latitude of the location at which the crime occurred.
Longitude: the longitude of the location at which the crime occurred.

### Problem 1.1 - Loading the Data

Read the dataset mvtWeek1.csv into R, using the read.csv function, and call the data frame "mvt". Remember to navigate to the directory on your computer containing the file mvtWeek1.csv first. It may take a few minutes to read in the data, since it is pretty large. Then, use the str and summary functions to answer the following questions.

### Problem 1.2 - Loading the Data
1 point possible (graded)
How many variables are in this dataset?
```{r}
mvt = read.csv("mvtWeek1.csv")
str(mvt)
summary(mvt)
```

### Problem 1.3 - Loading the Data
1 point possible (graded)
Using the "max" function, what is the maximum value of the variable "ID"?

### Problem 1.4 - Loading the Data
1 point possible (graded)
What is the minimum value of the variable "Beat"?
```{r}
max(mvt$ID)
min(mvt$Beat)
```




###Problem 1.5 - Loading the Data
1 point possible (graded)
How many observations have value TRUE in the Arrest variable (this is the number of crimes for which an arrest was made


###Problem 1.6 - Loading the Data
1 point possible (graded)
How many observations have a LocationDescription value of ALLEY?

```{r}
str(mvt)
summary(mvt)
```


###Problem 2.1 - Understanding Dates in R
In many datasets, like this one, you have a date field. Unfortunately, R does not automatically recognize entries that look like dates. We need to use a function in R to extract the date and time. Take a look at the first entry of Date (remember to use square brackets when looking at a certain entry of a variable).

In what format are the entries in the variable Date?

```{r}
sample(mvt$Date, 100)
```


###Problem 2.2 - Understanding Dates in R
1 point possible (graded)
Now, let's convert these characters into a Date object in R. In your R console, type

DateConvert = as.Date(strptime(mvt$Date, "%m/%d/%y %H:%M"))

This converts the variable "Date" into a Date object in R. Take a look at the variable DateConvert using the summary function.

What is the month and year of the median date in our dataset? Enter your answer as "Month Year", without the quotes. (Ex: if the answer was 2008-03-28, you would give the answer "March 2008", without the quotes.)



```{r}
DateConvert = as.Date(strptime(mvt$Date, "%m/%d/%y %H:%M"))
str(DateConvert)
summary(DateConvert)
```


###Problem 2.3 - Understanding Dates in R
1 point possible (graded)
Now, let's extract the month and the day of the week, and add these variables to our data frame mvt. We can do this with two simple functions. Type the following commands in R:

```{r}
mvt$Month = months(DateConvert)

mvt$Weekday = weekdays(DateConvert)
```


This creates two new variables in our data frame, Month and Weekday, and sets them equal to the month and weekday values that we can extract from the Date object. Lastly, replace the old Date variable with DateConvert by typing:


```{r}
mvt$Date = DateConvert
```

Using the table command, answer the following questions.

In which month did the fewest motor vehicle thefts occur?

###Problem 2.4 - Understanding Dates in R
1 point possible (graded)
On which weekday did the most motor vehicle thefts occur?

###Problem 2.5 - Understanding Dates in R
1 point possible (graded)
Each observation in the dataset represents a motor vehicle theft, and the Arrest variable indicates whether an arrest was later made for this theft. Which month has the largest number of motor vehicle thefts for which an arrest was made?


```{r}
which.min(table(mvt$Month))
which.max(table(mvt$Weekday))
sort((table(mvt$Month, mvt$Arrest)[,2]))

```

###Problem 3.1 - Visualizing Crime Trends
3 points possible (graded)
Now, let's make some plots to help us better understand how crime has changed over time in Chicago. Throughout this problem, and in general, you can save your plot to a file. For more information, this website very clearly explains the process.

First, let's make a histogram of the variable Date. We'll add an extra argument, to specify the number of bars we want in our histogram. In your R console, type

hist(mvt$Date, breaks=100)

Looking at the histogram, answer the following questions.

In general, does it look like crime increases or decreases from 2002 - 2012?

In general, does it look like crime increases or decreases from 2005 - 2008?


```{r}
hist(mvt$Date, breaks=100)

```



###Problem 3.2 - Visualizing Crime Trends
1/1 point (graded)
Now, let's see how arrests have changed over time. Create a boxplot of the variable "Date", sorted by the variable "Arrest" (if you are not familiar with boxplots and would like to learn more, check out this tutorial). In a boxplot, the bold horizontal line is the median value of the data, the box shows the range of values between the first quartile and third quartile, and the whiskers (the dotted lines extending outside the box) show the minimum and maximum values, excluding any outliers (which are plotted as circles). Outliers are defined by first computing the difference between the first and third quartile values, or the height of the box. This number is called the Inter-Quartile Range (IQR). Any point that is greater than the third quartile plus the IQR or less than the first quartile minus the IQR is considered an outlier.

Does it look like there were more crimes for which arrests were made in the first half of the time period or the second half of the time period? (Note that the time period is from 2001 to 2012, so the middle of the time period is the beginning of 2007.)

```{r}
boxplot(mvt$Date[mvt$Arrest==1])
```


###Problem 3.3 - Visualizing Crime Trends
2.0 points possible (graded)
Let's investigate this further. Use the table function for the next few questions.

For what proportion of motor vehicle thefts in 2001 was an arrest made?

Note: in this question and many others in the course, we are asking for an answer as a proportion. Therefore, your answer should take a value between 0 and 1.


###Problem 3.4 - Visualizing Crime Trends
1 point possible (graded)
For what proportion of motor vehicle thefts in 2007 was an arrest made?

###Problem 3.5 - Visualizing Crime Trends
1 point possible (graded)
For what proportion of motor vehicle thefts in 2012 was an arrest made?



Since there may still be open investigations for recent crimes, this could explain the trend we are seeing in the data. There could also be other factors at play, and this trend should be investigated further. However, since we don't know when the arrests were actually made, our detective work in this area has reached a dead end.
```{r}



#2001 arrest proportion
prop.table(table(mvt$Arrest[mvt$Year == "2001"]))

#2007 arrest proportion
prop.table(table(mvt$Arrest[mvt$Year == "2007"]))

#2012 arrest proportion
prop.table(table(mvt$Arrest[mvt$Year == "2012"]))

```

###Problem 4.1 - Popular Locations

Analyzing this data could be useful to the Chicago Police Department when deciding where to allocate resources. If they want to increase the number of arrests that are made for motor vehicle thefts, where should they focus their efforts?

We want to find the top five locations where motor vehicle thefts occur. If you create a table of the LocationDescription variable, it is unfortunately very hard to read since there are 78 different locations in the data set. By using the sort function, we can view this same table, but sorted by the number of observations in each category. In your R console, type:

sort(table(mvt$LocationDescription))

Which locations are the top five locations for motor vehicle thefts, excluding the "Other" category? You should select 5 of the following options.



```{r}
#Popular locations
sort(table(mvt$LocationDescription), decreasing = T)
```

###Problem 4.2 - Popular Locations
1 point possible (graded)
Create a subset of your data, only taking observations for which the theft happened in one of these five locations, and call this new data set "Top5". To do this, you can use the | symbol. In lecture, we used the & symbol to use two criteria to make a subset of the data. To only take observations that have a certain value in one variable or the other, the | character can be used in place of the & symbol. This is also called a logical "or" operation.

Alternately, you could create five different subsets, and then merge them together into one data frame using rbind.

How many observations are in Top5?


```{r}

#Top5 locations
Top5 = subset(mvt, LocationDescription == "STREET" | LocationDescription == "PARKING LOT/GARAGE(NON.RESID.)" | LocationDescription == "ALLEY"| LocationDescription == "GAS STATION" | LocationDescription == "DRIVEWAY - RESIDENTIAL")
str(Top5)



```

###Problem 4.3 - Popular Locations
2.0 points possible (graded)
R will remember the other categories of the LocationDescription variable from the original dataset, so running table(Top5$LocationDescription) will have a lot of unnecessary output. To make our tables a bit nicer to read, we can refresh this factor variable. In your R console, type:

```{r}
Top5$LocationDescription = factor(Top5$LocationDescription)

```

If you run the str or table function on Top5 now, you should see that LocationDescription now only has 5 values, as we expect.
```{r}
str(Top5)
```


Use the Top5 data frame to answer the remaining questions.

One of the locations has a much higher arrest rate than the other locations. Which is it? Please enter the text in exactly the same way as how it looks in the answer options for Problem 4.1.

```{r}
require(data.table)
TopDT <- data.table(Top5)
TopDT[, .(Amount = .N), .(LocationDescription, Arrest)][, .(Amount/sum(Amount), Arrest), .(LocationDescription)][order(-Arrest, -V1)]



```

###Problem 4.4 - Popular Locations
1 point possible (graded)
On which day of the week do the most motor vehicle thefts at gas stations happen?
```{r}
TopDT[LocationDescription == "GAS STATION", .N, Weekday][order(-N)]
```

###Problem 4.5 - Popular Locations
1 point possible (graded)
On which day of the week do the fewest motor vehicle thefts in residential driveways happen?

```{r}
TopDT[LocationDescription == "DRIVEWAY - RESIDENTIAL", .N, Weekday][order(N)]
```







How many rows of data (observations) are in this dataset?
## Assignment II STOCK DYNAMICS

A stock market is where buyers and sellers trade shares of a company, and is one of the most popular ways for individuals and companies to invest money. The size of the world stock market  is now estimated to be in the trillions. The largest stock market in the world is the New York Stock Exchange (NYSE), located in New York City. About 2,800 companies are listed on the NSYE. In this problem, we'll look at the monthly stock prices of five of these companies: IBM, General Electric (GE), Procter and Gamble, Coca Cola, and Boeing. The data used in this problem comes from Infochimps.

Download and read the following files into R, using the read.csv function: IBMStock.csv, GEStock.csv, ProcterGambleStock.csv, CocaColaStock.csv, and BoeingStock.csv. (Do not open these files in any spreadsheet software before completing this problem because it might change the format of the Date field.)

Call the data frames "IBM", "GE", "ProcterGamble", "CocaCola", and "Boeing", respectively. Each data frame has two variables, described as follows:

Date: the date of the stock price, always given as the first of the month.
StockPrice: the average stock price of the company in the given month.
In this problem, we'll take a look at how the stock dynamics of these companies have changed over time.



```{r}

IBM = read.csv("IBMStock.csv")
GE = read.csv("GEStock.csv")
ProcterGamble = read.csv("ProcterGambleStock.csv")
CocaCola = read.csv("CocaColaStock.csv")
Boeing = read.csv("BoeingStock.csv")

```

### Problem 1.1 - Summary Statistics

(1 point possible)
Before working with these data sets, we need to convert the dates into a format that R can understand. Take a look at the structure of one of the datasets using the str function. Right now, the date variable is stored as a factor. We can convert this to a "Date" object in R by using the following five commands (one for each data set):


```{r}
IBM$Date = as.Date(IBM$Date, "%m/%d/%y")

GE$Date = as.Date(GE$Date, "%m/%d/%y")

CocaCola$Date = as.Date(CocaCola$Date, "%m/%d/%y")

ProcterGamble$Date = as.Date(ProcterGamble$Date, "%m/%d/%y")

Boeing$Date = as.Date(Boeing$Date, "%m/%d/%y")

```



The first argument to the as.Date function is the variable we want to convert, and the second argument is the format of the Date variable. We can just overwrite the original Date variable values with the output of this function. Now, answer the following questions using the str and summary functions.

Our five datasets all have the same number of observations. How many observations are there in each data set?


```{r}
nrow(IBM)
```

### Problem 1.2 - Summary Statistics
What is the earliest year in our datasets?
```{r}
str(IBM)
str(GE)
str(CocaCola)
str(ProcterGamble)
str(Boeing)
```

### Problem 1.3 - 1.7 Summary Statistics

What is the latest year in our datasets?
What is the mean stock price of IBM over this time period?
What is the minimum stock price of General Electric (GE) over this time period?
What is the median stock price of Boeing over this time period?

```{r}
summary(IBM)
summary(GE)
summary(CocaCola)
summary(ProcterGamble)
summary(Boeing)
```

What is the standard deviation of the stock price of Procter & Gamble over this time period?

```{r}

sd(ProcterGamble$StockPrice)

```

### Problem 2.1 - Visualizing Stock Dynamics

Let's plot the stock prices to see if we can visualize trends in stock prices during this time period. Using the plot function, plot the Date on the x-axis and the StockPrice on the y-axis, for Coca-Cola.

This plots our observations as points, but we would really like to see a line instead, since this is a continuous time period. To do this, add the argument type="l" to your plot command, and re-generate the plot (the character is quotes is the letter l, for line). You should now see a line plot of the Coca-Cola stock price.

Around what year did Coca-Cola has its highest stock price in this time period?
```{r}

plot(CocaCola, type = "l")

```


### Problem 2.2 - Visualizing Stock Dynamics

Now, let's add the line for Procter & Gamble too. You can add a line to a plot in R by using the lines function instead of the plot function. Keeping the plot for Coca-Cola open, type in your R console:

lines(ProcterGamble$Date, ProcterGamble$StockPrice)

Unfortunately, it's hard to tell which line is which. Let's fix this by giving each line a color. First, re-run the plot command for Coca-Cola, but add the argument col="red". You should see the plot for Coca-Cola show up again, but this time in red. Now, let's add the Procter & Gamble line (using the lines function like we did before), adding the argument col="blue". You should now see in your plot the Coca-Cola stock price in red, and the Procter & Gamble stock price in blue.

As an alternative choice to changing the colors, you could instead change the line type of the Procter & Gamble line by adding the argument lty=2. This will make the Procter & Gamble line dashed.

Using this plot, answer the following questions.

In March of 2000, the technology bubble burst, and a stock market crash occurred. According to this plot, which company's stock dropped more?


To answer this question and the ones that follow, you may find it useful to draw a vertical line at a certain date. To do this, type the command

abline(v=as.Date(c("2000-03-01")), lwd=2)

in your R console, with the plot still open. This generates a vertical line at the date March 1, 2000. The argument lwd=2 makes the line a little thicker. You can change the date in this command to generate the vertical line in different locations.

```{r}

plot(CocaCola, type = "l", col = "red")
lines(ProcterGamble$Date, ProcterGamble$StockPrice, col = "blue")
abline(v=as.Date(c("2000-03-01")), lwd=2)
```

### Problem 2.3 - Visualizing Stock Dynamics

Answer these questions using the plot you generated in the previous problem.

Around 1983, the stock for one of these companies (Coca-Cola or Procter and Gamble) was going up, while the other was going down. Which one was going up?

```{r}

plot(CocaCola, type = "l", col = "red")
lines(ProcterGamble$Date, ProcterGamble$StockPrice, col = "blue")
abline(v=as.Date(c("1983-01-01")), lwd=2)
```


### Problem 3.1 - Visualizing Stock Dynamics 1995-2005

(1 point possible)
Let's take a look at how the stock prices changed from 1995-2005 for all five companies. In your R console, start by typing the following plot command:

plot(CocaCola$Date[301:432], CocaCola$StockPrice[301:432], type="l", col="red", ylim=c(0,210))

This will plot the CocaCola stock prices from 1995 through 2005, which are the observations numbered from 301 to 432. The additional argument, ylim=c(0,210), makes the y-axis range from 0 to 210. This will allow us to see all of the stock values when we add in the other companies.

Now, use the lines function to add in the other four companies, remembering to only plot the observations from 1995 to 2005, or [301:432]. You don't need the "type" or "ylim" arguments for the lines function, but remember to make each company a different color so that you can tell them apart. Some color options are "red", "blue", "green", "purple", "orange", and "black". To see all of the color options in R, type colors() in your R console.

(If you prefer to change the type of the line instead of the color, here are some options for changing the line type: lty=2 will make the line dashed, lty=3 will make the line dotted, lty=4 will make the line alternate between dashes and dots, and lty=5 will make the line long-dashed.)

Use this plot to answer the following four questions.

Which stock fell the most right after the technology bubble burst in March 2000?

```{r}

require(ggplot2)
ggplot(data = CocaCola, aes(x = Date, y = StockPrice)) + geom_line(data = CocaCola[301:432,], aes (colour = "CocaCola")) + geom_line(data = ProcterGamble[301:432,], aes (colour = "ProcterGamble")) + geom_line(data = GE[301:432,], aes (colour = "GE")) + geom_line(data = IBM[301:432,], aes (colour = "IBM")) + geom_line(data = Boeing[301:432,], aes (colour = "Boeing")) + scale_color_manual("",breaks = c("CocaCola","ProcterGamble","GE","IBM","Boeing"), values = c("blue", "red","green","skyblue","tan")) + geom_vline(xintercept = as.numeric(as.Date("2000-03-15")))

```

### Problem 3.2 - Visualizing Stock Dynamics 1995-2005


Which stock reaches the highest value in the time period 1995-2005?

```{r}

ggplot(data = CocaCola, aes(x = Date, y = StockPrice)) + geom_line(data = CocaCola[301:432,], aes (colour = "CocaCola")) + geom_line(data = ProcterGamble[301:432,], aes (colour = "ProcterGamble")) + geom_line(data = GE[301:432,], aes (colour = "GE")) + geom_line(data = IBM[301:432,], aes (colour = "IBM")) + geom_line(data = Boeing[301:432,], aes (colour = "Boeing")) + scale_color_manual("",breaks = c("CocaCola","ProcterGamble","GE","IBM","Boeing"), values = c("blue", "red","green","skyblue","tan")) + geom_vline(xintercept = as.numeric(as.Date("1997-09-01"))) + geom_vline(xintercept = as.numeric(as.Date("1997-11-30")))
```


### Problem 3.3 - Visualizing Stock Dynamics 1995-2005


In October of 1997, there was a global stock market crash that was caused by an economic crisis in Asia. Comparing September 1997 to November 1997, which companies saw a decreasing trend in their stock price? (Select all that apply.)

```{r}
ggplot(data = CocaCola, aes(x = Date, y = StockPrice)) + geom_line(data = CocaCola[301:432,], aes (colour = "CocaCola")) + geom_line(data = ProcterGamble[301:432,], aes (colour = "ProcterGamble")) + geom_line(data = GE[301:432,], aes (colour = "GE")) + geom_line(data = IBM[301:432,], aes (colour = "IBM")) + geom_line(data = Boeing[301:432,], aes (colour = "Boeing")) + scale_color_manual("",breaks = c("CocaCola","ProcterGamble","GE","IBM","Boeing"), values = c("blue", "red","green","skyblue","tan")) + geom_vline(xintercept = as.numeric(as.Date("1997-11-30"))) + geom_vline(xintercept = as.numeric(as.Date("1997-09-01")))
```


### Problem 3.4 - Visualizing Stock Dynamics 1995-2005
ry

In the last two years of this time period (2004 and 2005) which stock seems to be performing the best, in terms of increasing stock price?

```{r}
ggplot(data = CocaCola, aes(x = Date, y = StockPrice)) + geom_line(data = CocaCola[301:432,], aes (colour = "CocaCola")) + geom_line(data = ProcterGamble[301:432,], aes (colour = "ProcterGamble")) + geom_line(data = GE[301:432,], aes (colour = "GE")) + geom_line(data = IBM[301:432,], aes (colour = "IBM")) + geom_line(data = Boeing[301:432,], aes (colour = "Boeing")) + scale_color_manual("",breaks = c("CocaCola","ProcterGamble","GE","IBM","Boeing"), values = c("blue", "red","green","skyblue","tan")) + geom_vline(xintercept = as.numeric(as.Date("2004-01-01"))) + geom_vline(xintercept = as.numeric(as.Date("2005-12-31")))
```



### Problem 4.1 - Monthly Trends

(1 point possible)
Lastly, let's see if stocks tend to be higher or lower during certain months. Use the tapply command to calculate the mean stock price of IBM, sorted by months. To sort by months, use

months(IBM$Date)

as the second argument of the tapply function.

For IBM, compare the monthly averages to the overall average stock price. In which months has IBM historically had a higher stock price (on average)? Select all that apply.

```{r}
monthPriceIBM <- tapply(IBM$StockPrice, months(IBM$Date), mean)/mean((IBM$StockPrice))
monthPriceIBM[monthPriceIBM>1]
```


###Problem 4.2 - Monthly Trends

(1 point possible)
Repeat the tapply function from the previous problem for each of the other four companies, and use the output to answer the remaining questions.

General Electric and Coca-Cola both have their highest average stock price in the same month. Which month is this?

###Problem 4.3 - Monthly Trends
1 point possible (graded)
For the months of December and January, every company's average stock is higher in one month and lower in the other. In which month are the stock prices lower?
```{r}
monthPriceGE <- tapply(GE$StockPrice, months(GE$Date), mean)/mean((GE$StockPrice))
monthPriceGE
monthPriceGE[which.max(monthPriceGE)]

monthPriceCocaCola <- tapply(CocaCola$StockPrice, months(CocaCola$Date), mean)/mean((CocaCola$StockPrice))
monthPriceCocaCola
monthPriceCocaCola[which.max(monthPriceCocaCola)]

```






## Assingment III: DEMOGRAPHICS AND EMPLOYMENT IN THE UNITED STATES

In the wake of the Great Recession of 2009, there has been a good deal of focus on employment statistics, one of the most important metrics policymakers use to gauge the overall strength of the economy. In the United States, the government measures unemployment using the Current Population Survey (CPS), which collects demographic and employment information from a wide range of Americans each month. In this exercise, we will employ the topics reviewed in the lectures as well as a few new techniques using the September 2013 version of this rich, nationally representative dataset (available online).

The observations in the dataset represent people surveyed in the September 2013 CPS who actually completed a survey. While the full dataset has 385 variables, in this exercise we will use a more compact version of the dataset, CPSData.csv, which has the following variables:

PeopleInHousehold: The number of people in the interviewee's household.

Region: The census region where the interviewee lives.

State: The state where the interviewee lives.

MetroAreaCode: A code that identifies the metropolitan area in which the interviewee lives (missing if the interviewee does not live in a metropolitan area). The mapping from codes to names of metropolitan areas is provided in the file MetroAreaCodes.csv.

Age: The age, in years, of the interviewee. 80 represents people aged 80-84, and 85 represents people aged 85 and higher.

Married: The marriage status of the interviewee.

Sex: The sex of the interviewee.

Education: The maximum level of education obtained by the interviewee.

Race: The race of the interviewee.

Hispanic: Whether the interviewee is of Hispanic ethnicity.

CountryOfBirthCode: A code identifying the country of birth of the interviewee. The mapping from codes to names of countries is provided in the file CountryCodes.csv.

Citizenship: The United States citizenship status of the interviewee.

EmploymentStatus: The status of employment of the interviewee.

Industry: The industry of employment of the interviewee (only available if they are employed).

###Problem 1.1 - Loading and Summarizing the Dataset

(1 point possible)
Load the dataset from CPSData.csv into a data frame called CPS, and view the dataset with the summary() and str() commands.

How many interviewees are in the dataset?

###Problem 1.2 - Loading and Summarizing the Dataset


Among the interviewees with a value reported for the Industry variable, what is the most common industry of employment? Please enter the name exactly how you see it.


```{r}
#Again, let's start by emptying the workspace:
rm (list = ls(all=T))
``` 


```{r}
CPS = read.csv("CPSData.csv")
str(CPS)
summary(CPS)
``` 




###Problem 1.3 - Loading and Summarizing the Dataset

Recall from the homework assignment "The Analytical Detective" that you can call the sort() function on the output of the table() function to obtain a sorted breakdown of a variable. For instance, sort(table(CPS$Region)) sorts the regions by the number of interviewees from that region.

Which state has the fewest interviewees?

```{r}
sort(table(CPS$State))
``` 




###Problem 1.4 - Loading and Summarizing the Dataset

(1 point possible)
What proportion of interviewees are citizens of the United States?
```{r}
(table(CPS$Citizenship)["Citizen, Native"]+table(CPS$Citizenship)["Citizen, Naturalized"])/sum(table(CPS$Citizenship))
```


###Problem 1.5 - Loading and Summarizing the Dataset

(1/1 point)
The CPS differentiates between race (with possible values American Indian, Asian, Black, Pacific Islander, White, or Multiracial) and ethnicity. A number of interviewees are of Hispanic ethnicity, as captured by the Hispanic variable. For which races are there at least 250 interviewees in the CPS dataset of Hispanic ethnicity? (Select all that apply.)

```{r}
table(CPS$Race[(CPS$Hispanic) == 1])

```



###Problem 2.1 - Evaluating Missing Values


Which variables have at least one interviewee with a missing (NA) value? (Select all that apply.)
```{r}
apply(CPS, 2, anyNA)
?apply
```


Problem 2.2 - Evaluating Missing Values

(1 point possible)
Often when evaluating a new dataset, we try to identify if there is a pattern in the missing values in the dataset. We will try to determine if there is a pattern in the missing values of the Married variable. The function 
```{r}
is.na(CPS$Married)
```
returns a vector of TRUE/FALSE values for whether the Married variable is missing. We can see the breakdown of whether Married is missing based on the reported value of the Region variable with the function
```{r}
table(CPS$Region, is.na(CPS$Married))
```

Which is the most accurate:



```{r}
table(CPS$Region, is.na(CPS$Married))
table(CPS$Sex, is.na(CPS$Married))
table(CPS$Age, is.na(CPS$Married))
table(CPS$Citizenship, is.na(CPS$Married))

```


###Problem 2.3 - Evaluating Missing Values

As mentioned in the variable descriptions, MetroAreaCode is missing if an interviewee does not live in a metropolitan area. Using the same technique as in the previous question, answer the following questions about people who live in non-metropolitan areas.

How many states had all interviewees living in a non-metropolitan area (aka they have a missing MetroAreaCode value)? For this question, treat the District of Columbia as a state (even though it is not technically a state).

```{r}
sum(table(CPS$State, is.na(CPS$MetroAreaCode))[,1] == 0)


sum(table(CPS$State, is.na(CPS$MetroAreaCode))[,2] == 0)

```



###Problem 2.4 - Evaluating Missing Values

Which region of the United States has the largest proportion of interviewees living in a non-metropolitan area?
```{r}
sort(prop.table(table(CPS$Region, is.na(CPS$MetroAreaCode)),1)[,2], decreasing = TRUE)
```


###Problem 2.5 - Evaluating Missing Values

While we were able to use the table() command to compute the proportion of interviewees from each region not living in a metropolitan area, it was somewhat tedious (it involved manually computing the proportion for each region) and isn't something you would want to do if there were a larger number of options. It turns out there is a less tedious way to compute the proportion of values that are TRUE. The mean() function, which takes the average of the values passed to it, will treat TRUE as 1 and FALSE as 0, meaning it returns the proportion of values that are true. For instance, mean(c(TRUE, FALSE, TRUE, TRUE)) returns 0.75. Knowing this, use tapply() with the mean function to answer the following questions:

Which state has a proportion of interviewees living in a non-metropolitan area closest to 30%?

Which state has the largest proportion of non-metropolitan interviewees, ignoring states where all interviewees were non-metropolitan?

```{r}
sort(prop.table(table(CPS$State, is.na(CPS$MetroAreaCode)),1)[,2], decreasing = TRUE)
```

###Problem 3.1 - Integrating Metropolitan Area Data

Codes like MetroAreaCode and CountryOfBirthCode are a compact way to encode factor variables with text as their possible values, and they are therefore quite common in survey datasets. In fact, all but one of the variables in this dataset were actually stored by a numeric code in the original CPS datafile.

When analyzing a variable stored by a numeric code, we will often want to convert it into the values the codes represent. To do this, we will use a dictionary, which maps the the code to the actual value of the variable. We have provided dictionaries MetroAreaCodes.csv and CountryCodes.csv, which respectively map MetroAreaCode and CountryOfBirthCode into their true values. Read these two dictionaries into data frames MetroAreaMap and CountryMap.

How many observations (codes for metropolitan areas) are there in MetroAreaMap?

```{r}
MetroAreaMap = read.csv("MetroAreaCodes.csv")
CountryMap = read.csv("CountryCodes.csv")
str(MetroAreaMap)
str(CountryMap)
```


###Problem 3.2 - Integrating Metropolitan Area Data

To merge in the metropolitan areas, we want to connect the field MetroAreaCode from the CPS data frame with the field Code in MetroAreaMap. The following command merges the two data frames on these columns, overwriting the CPS data frame with the result:

CPS = merge(CPS, MetroAreaMap, by.x="MetroAreaCode", by.y="Code", all.x=TRUE)

The first two arguments determine the data frames to be merged (they are called "x" and "y", respectively, in the subsequent parameters to the merge function). by.x="MetroAreaCode" means we're matching on the MetroAreaCode variable from the "x" data frame (CPS), while by.y="Code" means we're matching on the Code variable from the "y" data frame (MetroAreaMap). Finally, all.x=TRUE means we want to keep all rows from the "x" data frame (CPS), even if some of the rows' MetroAreaCode doesn't match any codes in MetroAreaMap (for those familiar with database terminology, this parameter makes the operation a left outer join instead of an inner join).

Review the new version of the CPS data frame with the summary() and str() functions. What is the name of the variable that was added to the data frame by the merge() operation?


```{r}
CPS = merge(CPS, MetroAreaMap, by.x="MetroAreaCode", by.y="Code", all.x=TRUE)
summary(CPS)
str(CPS)
```

###Problem 3.3 - Integrating Metropolitan Area Data

Which of the following metropolitan areas has the largest number of interviewees?

```{r}
head(sort(table(subset(CPS, MetroArea == "Atlanta-Sandy Springs-Marietta, GA" | MetroArea == "Baltimore-Towson, MD" | MetroArea == "Boston-Cambridge-Quincy, MA-NH" | MetroArea == "San Francisco-Oakland-Fremont, CA", select =c(MetroArea))), decreasing = TRUE))
```


###Problem 3.4 - Integrating Metropolitan Area Data

Which metropolitan area has the highest proportion of interviewees of Hispanic ethnicity? Hint: Use tapply() with mean, as in the previous subproblem. Calling sort() on the output of tapply() could also be helpful here.

```{r}
sort(tapply(CPS$Hispanic, CPS$MetroArea, mean), decreasing = T)[1]
```

###Problem 3.5 - Integrating Metropolitan Area Data
2.0 points possible (graded)
Remembering that CPS$Race == "Asian" returns a TRUE/FALSE vector of whether an interviewee is Asian, determine the number of metropolitan areas in the United States from which at least 20% of interviewees are Asian.

```{r}
sort(tapply(CPS$Race=="Asian", CPS$MetroArea, mean, na.rm=T))
```



###Problem 3.6 - Integrating Metropolitan Area Data

Normally, we would look at the sorted proportion of interviewees from each metropolitan area who have not received a high school diploma with the command:

```{r}
sort(tapply(CPS$Education == "No high school diploma", CPS$MetroArea, mean))
```


However, none of the interviewees aged 14 and younger have an education value reported, so the mean value is reported as NA for each metropolitan area. To get mean (and related functions, like sum) to ignore missing values, you can pass the parameter na.rm=TRUE. Passing na.rm=TRUE to the tapply function, determine which metropolitan area has the smallest proportion of interviewees who have received no high school diploma.

```{r}
sort(tapply(CPS$Education == "No high school diploma", CPS$MetroArea, mean, na.rm = T))
```

###Problem 4.1 - Integrating Country of Birth Data

Just as we did with the metropolitan area information, merge in the country of birth information from the CountryMap data frame, replacing the CPS data frame with the result. If you accidentally overwrite CPS with the wrong values, remember that you can restore it by re-loading the data frame from CPSData.csv and then merging in the metropolitan area information using the command provided in the previous subproblem.

What is the name of the variable added to the CPS data frame by this merge operation?

How many interviewees have a missing value for the new country of birth variable?

```{r}
str(CPS)
str(CountryMap)
summary(CPS)
CPS = merge(CPS, CountryMap, by.x="CountryOfBirthCode", by.y="Code", all.x=TRUE)
summary(CPS)
sort(table(CPS$Country))
```


###Problem 4.3 - Integrating Country of Birth Data

What proportion of the interviewees from the "New York-Northern New Jersey-Long Island, NY-NJ-PA" metropolitan area have a country of birth that is not the United States? For this computation, don't include people from this metropolitan area who have a missing country of birth.

```{r}
CPSDT <- data.table(CPS)
CPSDT
CPSDT[MetroArea == "New York-Northern New Jersey-Long Island, NY-NJ-PA", .N, .(Country)][,.(Country, 1-(N/sum(N)))] #wanted to do it in one line, although this is really, really inelegant
```


Which metropolitan area has the largest number (note -- not proportion) of interviewees with a country of birth in India? Hint -- remember to include na.rm=TRUE if you are using tapply() to answer this question.

In Brazil?

In Somalia?

```{r}
CPSDT[Country == "India", .N, MetroArea][order(N)]
CPSDT[Country == "Brazil", .N, MetroArea][order(N)]
CPSDT[Country == "Somalia", .N, MetroArea][order(N)]
```

