---
title: "Homework_w2"
author: "Joona Rauhamäki"
date: "30. April 2016"
output: html_document
---

# The Analytics Edget Homework week 2

## CLIMATE CHANGE

There have been many studies documenting that the average global temperature has been increasing over the last century. The consequences of a continued rise in global temperature will be dire. Rising sea levels and an increased frequency of extreme weather events will affect billions of people.

In this problem, we will attempt to study the relationship between average global temperature and several other factors.

The file climate_change.csv contains climate data from May 1983 to December 2008. The available variables include:

Year: the observation year.
Month: the observation month.
Temp: the difference in degrees Celsius between the average global temperature in that period and a reference value. This data comes from the Climatic Research Unit at the University of East Anglia.
CO2, N2O, CH4, CFC.11, CFC.12: atmospheric concentrations of carbon dioxide (CO2), nitrous oxide (N2O), methane  (CH4), trichlorofluoromethane (CCl3F; commonly referred to as CFC-11) and dichlorodifluoromethane (CCl2F2; commonly referred to as CFC-12), respectively. This data comes from the ESRL/NOAA Global Monitoring Division.
CO2, N2O and CH4 are expressed in ppmv (parts per million by volume  -- i.e., 397 ppmv of CO2 means that CO2 constitutes 397 millionths of the total volume of the atmosphere)
CFC.11 and CFC.12 are expressed in ppbv (parts per billion by volume). 
Aerosols: the mean stratospheric aerosol optical depth at 550 nm. This variable is linked to volcanoes, as volcanic eruptions result in new particles being added to the atmosphere, which affect how much of the sun's energy is reflected back into space. This data is from the Godard Institute for Space Studies at NASA.
TSI: the total solar irradiance (TSI) in W/m2 (the rate at which the sun's energy is deposited per unit area). Due to sunspots and other solar phenomena, the amount of energy that is given off by the sun varies substantially with time. This data is from the SOLARIS-HEPPA project website.
MEI: multivariate El Nino Southern Oscillation index (MEI), a measure of the strength of the El Nino/La Nina-Southern Oscillation (a weather effect in the Pacific Ocean that affects global temperatures). This data comes from the ESRL/NOAA Physical Sciences Division.

```{r}
#start by emptying the environment
rm (list = ls(all=T))
```

## Problem 1.1 - Creating Our First Model

We are interested in how changes in these variables affect future temperatures, as well as how well these variables explain temperature changes so far. To do this, first read the dataset climate_change.csv into R.

```{r}
#read data into R
climateFull = read.csv("climate_change.csv")
str(climateFull)
summary(climateFull)
```

Then, split the data into a training set, consisting of all the observations up to and including 2006, and a testing set consisting of the remaining years (hint: use subset). A training set refers to the data that will be used to build the model (this is the data we give to the lm() function), and a testing set refers to the data we will use to test our predictive ability.

```{r}
climateTrain = subset(climateFull, Year <= 2006)
summary(climateTrain)
climateTest = subset(climateFull, Year > 2006)
summary(climateTest)
```


Next, build a linear regression model to predict the dependent variable Temp, using MEI, CO2, CH4, N2O, CFC.11, CFC.12, TSI, and Aerosols as independent variables (Year and Month should NOT be used in the model). Use the training set to build the model.

Enter the model R2 (the "Multiple R-squared" value):

```{r}
model1 = lm(Temp ~ . -Year -Month, data = climateTrain)
summary(model1)
```

## Problem 1.2 - Creating Our First Model

Which variables are significant in the model? We will consider a variable signficant only if the p-value is below 0.05. (Select all that apply.)

## Problem 2.1 - Understanding the Model


Current scientific opinion is that nitrous oxide and CFC-11 are greenhouse gases: gases that are able to trap heat from the sun and contribute to the heating of the Earth. However, the regression coefficients of both the N2O and CFC-11 variables are negative, indicating that increasing atmospheric concentrations of either of these two compounds is associated with lower global temperatures.

Which of the following is the simplest correct explanation for this contradiction?

##Problem 2.2 - Understanding the Model


Compute the correlations between all the variables in the training set. Which of the following independent variables is N2O highly correlated with (absolute correlation greater than 0.7)? Select all that apply.

Which of the following independent variables is CFC.11 highly correlated with? Select all that apply.

```{r}
cor(climateTrain)
```

##Problem 3 - Simplifying the Model

Given that the correlations are so high, let us focus on the N2O variable and build a model with only MEI, TSI, Aerosols and N2O as independent variables. Remember to use the training set to build the model.
```{r}
model2 = lm(Temp ~ MEI+TSI+Aerosols+N2O, data = climateTrain)
summary(model2)
```


Enter the coefficient of N2O in this reduced model:

(How does this compare to the coefficient in the previous model with all of the variables?)

Enter the model R2:


## Problem 4 - Automatically Building the Model

We have many variables in this problem, and as we have seen above, dropping some from the model does not decrease model quality. R provides a function, step, that will automate the procedure of trying different combinations of variables to find a good compromise of model simplicity and R2. This trade-off is formalized by the Akaike information criterion (AIC) - it can be informally thought of as the quality of the model with a penalty for the number of variables in the model.

The step function has one argument - the name of the initial model. It returns a simplified model. Use the step function in R to derive a new model, with the full model as the initial model (HINT: If your initial full model was called "climateLM", you could create a new model with the step function by typing step(climateLM). Be sure to save your new model to a variable name so that you can look at the summary. For more information about the step function, type ?step in your R console.)

```{r}
stepModel = step(model1)
summary(stepModel)
```

Enter the R2 value of the model produced by the step function:


## Problem 5 - Testing on Unseen Data

(2 points possible)
We have developed an understanding of how well we can fit a linear regression to the training data, but does the model quality hold when applied to unseen data?

Using the model produced from the step function, calculate temperature predictions for the testing data set, using the predict function.

```{r}
pred = predict(stepModel, newdata = climateTest)
SSE = sum((pred-climateTest$Temp)^2)
SST = sum((mean(climateTrain$Temp)-climateTest$Temp)^2)
R2 = 1-(SSE/SST)
R2
```

Enter the testing set R2:


## READING TEST SCORES

The Programme for International Student Assessment (PISA) is a test given every three years to 15-year-old students from around the world to evaluate their performance in mathematics, reading, and science. This test provides a quantitative way to compare the performance of students from different parts of the world. In this homework assignment, we will predict the reading scores of students from the United States of America on the 2009 PISA exam.

The datasets pisa2009train.csv and pisa2009test.csv contain information about the demographics and schools for American students taking the exam, derived from 2009 PISA Public-Use Data Files distributed by the United States National Center for Education Statistics (NCES). While the datasets are not supposed to contain identifying information about students taking the test, by using the data you are bound by the NCES data use agreement, which prohibits any attempt to determine the identity of any student in the datasets.

Each row in the datasets pisa2009train.csv and pisa2009test.csv represents one student taking the exam. The datasets have the following variables:

grade: The grade in school of the student (most 15-year-olds in America are in 10th grade)

male: Whether the student is male (1/0)

raceeth: The race/ethnicity composite of the student

preschool: Whether the student attended preschool (1/0)

expectBachelors: Whether the student expects to obtain a bachelor's degree (1/0)

motherHS: Whether the student's mother completed high school (1/0)

motherBachelors: Whether the student's mother obtained a bachelor's degree (1/0)

motherWork: Whether the student's mother has part-time or full-time work (1/0)

fatherHS: Whether the student's father completed high school (1/0)

fatherBachelors: Whether the student's father obtained a bachelor's degree (1/0)

fatherWork: Whether the student's father has part-time or full-time work (1/0)

selfBornUS: Whether the student was born in the United States of America (1/0)

motherBornUS: Whether the student's mother was born in the United States of America (1/0)

fatherBornUS: Whether the student's father was born in the United States of America (1/0)

englishAtHome: Whether the student speaks English at home (1/0)

computerForSchoolwork: Whether the student has access to a computer for schoolwork (1/0)

read30MinsADay: Whether the student reads for pleasure for 30 minutes/day (1/0)

minutesPerWeekEnglish: The number of minutes per week the student spend in English class

studentsInEnglish: The number of students in this student's English class at school

schoolHasLibrary: Whether this student's school has a library (1/0)

publicSchool: Whether this student attends a public school (1/0)

urban: Whether this student's school is in an urban area (1/0)

schoolSize: The number of students in this student's school

readingScore: The student's reading score, on a 1000-point scale

```{r}
#start by emptying the environment
rm (list = ls(all=T))
```


## Problem 1.1 - Dataset size

(1 point possible)
Load the training and testing sets using the read.csv() function, and save them as variables with the names pisaTrain and pisaTest.

How many students are there in the training set?

```{r}
pisaTrain = read.csv("pisa2009train.csv")
pisaTest = read.csv("pisa2009test.csv")
summary(pisaTrain)
str(pisaTrain)
```


## Problem 1.2 - Summarizing the dataset

Using tapply() on pisaTrain, what is the average reading test score of males?

```{r}
tapply(pisaTrain$readingScore, pisaTrain$male, mean)
```


## Problem 1.3 - Locating missing values


Which variables are missing data in at least one observation in the training set? Select all that apply.
```{r}
 apply(pisaTrain, 2, anyNA)
```


## Problem 1.4 - Removing missing values

Linear regression discards observations with missing data, so we will remove all such observations from the training and testing sets. Later in the course, we will learn about imputation, which deals with missing data by filling in missing values with plausible information.

Type the following commands into your R console to remove observations with any missing value from pisaTrain and pisaTest:

pisaTrain = na.omit(pisaTrain)

pisaTest = na.omit(pisaTest)
```{r}
pisaTrain = na.omit(pisaTrain)

pisaTest = na.omit(pisaTest)

str(pisaTrain)
str(pisaTest)
```
How many observations are now in the training set?


 - unanswered  
How many observations are now in the testing set?


## Problem 2.1 - Factor variables

(2/2 points)
Factor variables are variables that take on a discrete set of values, like the "Region" variable in the WHO dataset from the second lecture of Unit 1. This is an unordered factor because there isn't any natural ordering between the levels. An ordered factor has a natural ordering between the levels (an example would be the classifications "large," "medium," and "small").

Which of the following variables is an unordered factor with at least 3 levels? (Select all that apply.)




## Problem 2.2 - Unordered factors in regression models

To include unordered factors in a linear regression model, we define one level as the "reference level" and add a binary variable for each of the remaining levels. In this way, a factor with n levels is replaced by n-1 binary variables. The reference level is typically selected to be the most frequently occurring level in the dataset.

As an example, consider the unordered factor variable "color", with levels "red", "green", and "blue". If "green" were the reference level, then we would add binary variables "colorred" and "colorblue" to a linear regression problem. All red examples would have colorred=1 and colorblue=0. All blue examples would have colorred=0 and colorblue=1. All green examples would have colorred=0 and colorblue=0.

Now, consider the variable "raceeth" in our problem, which has levels "American Indian/Alaska Native", "Asian", "Black", "Hispanic", "More than one race", "Native Hawaiian/Other Pacific Islander", and "White". Because it is the most common in our population, we will select White as the reference level.

Which binary variables will be included in the regression model? (Select all that apply.)


## Problem 2.3 - Example unordered factors

Consider again adding our unordered factor race to the regression model with reference level "White".

For a student who is Asian, which binary variables would be set to 0? All remaining variables will be set to 1. (Select all that apply.)



## Problem 3.1 - Building a model

Because the race variable takes on text values, it was loaded as a factor variable when we read in the dataset with read.csv() -- you can see this when you run str(pisaTrain) or str(pisaTest). However, by default R selects the first level alphabetically ("American Indian/Alaska Native") as the reference level of our factor instead of the most common level ("White"). Set the reference level of the factor by typing the following two lines in your R console:
```{r}
pisaTrain$raceeth = relevel(pisaTrain$raceeth, "White")

pisaTest$raceeth = relevel(pisaTest$raceeth, "White")
```


Now, build a linear regression model (call it lmScore) using the training set to predict readingScore using all the remaining variables.

```{r}
lmScore = lm(readingScore ~ ., data = pisaTrain)
summary(lmScore)
```

It would be time-consuming to type all the variables, but R provides the shorthand notation "readingScore ~ ." to mean "predict readingScore using all the other variables in the data frame." The period is used to replace listing out all of the independent variables. As an example, if your dependent variable is called "Y", your independent variables are called "X1", "X2", and "X3", and your training data set is called "Train", instead of the regular notation:

LinReg = lm(Y ~ X1 + X2 + X3, data = Train)

You would use the following command to build your model:

LinReg = lm(Y ~ ., data = Train)

What is the Multiple R-squared value of lmScore on the training set?


## Problem 3.2 - Computing the root-mean squared error of the model

What is the training-set root-mean squared error (RMSE) of lmScore?


```{r}
SSE = sum((lmScore$residuals)^2)
RMSE = sqrt(SSE/nrow(pisaTrain))
RMSE
```



## Problem 3.3 - Comparing predictions for similar students

Consider two students A and B. They have all variable values the same, except that student A is in grade 11 and student B is in grade 9. What is the predicted reading score of student A minus the predicted reading score of student B?

## Problem 3.4 - Interpreting model coefficients

What is the meaning of the coefficient associated with variable raceethAsian?

## Problem 3.5 - Identifying variables lacking statistical significance

Based on the significance codes, which variables are candidates for removal from the model? Select all that apply. (We'll assume that the factor variable raceeth should only be removed if none of its levels are significant.)


##Problem 4.1 - Predicting on unseen data


Using the "predict" function and supplying the "newdata" argument, use the lmScore model to predict the reading scores of students in pisaTest. Call this vector of predictions "predTest". Do not change the variables in the model (for example, do not remove variables that we found were not significant in the previous part of this problem). Use the summary function to describe the test set predictions.

What is the range between the maximum and minimum predicted reading score on the test set?

```{r}
predTest = predict(lmScore, newdata = pisaTest)
summary(predTest)
max(predTest)-min(predTest)
```


## Problem 4.2 - Test set SSE and RMSE

What is the sum of squared errors (SSE) of lmScore on the testing set?

What is the root-mean squared error (RMSE) of lmScore on the testing set?

```{r}
SSE = sum((predTest - pisaTest$readingScore)^2)
SSE
RMSE = sqrt(SSE/nrow(pisaTest))
RMSE
```


##Problem 4.3 - Baseline prediction and test-set SSE


What is the predicted test score used in the baseline model? Remember to compute this value using the training set and not the test set.

What is the sum of squared errors of the baseline model on the testing set? HINT: We call the sum of squared errors for the baseline model the total sum of squares (SST).

```{r}
baseScore = mean(pisaTrain$readingScore)
baseScore
SST = sum((baseScore-pisaTest$readingScore)^2)
SST
```



Problem 4.4 - Test-set R-squared

What is the test-set R-squared value of lmScore?

```{r}
R2 = 1-(SSE/SST)
R2
```